\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-09-20 pkg/evtree/inst/doc/evtree.Rnw
improved intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@245 255f4249-2263-4417-b1ea-8498779e27a7

\Abstract{
  Commonly used classification and regression tree methods like the CART
  algorithm are recursive partitioning methods that build the model in a forward
  stepwise search. Although this approach is known to be an efficient heuristic,
  the results of recursive tree methods are only locally optimal, as splits are
  chosen to maximize homogeneity at the next step only. An alternative way to
  search over the parameter space of trees is to use global optimization methods
  like evolutionary algorithms. This paper describes the \pkg{evtree} package,
  which implements an evolutionary algorithm for learning globally optimal
  classification and regression trees in \proglang{R}. CPU and memory-intensive
  tasks are fully computed in \proglang{C++} while the \pkg{partykit}
  \citep{hothorn2009partykit} package is leveraged to represent the resulting
  trees in \proglang{R}, providing unified infrastructure for summaries,
  visualizations, and predictions. \code{evtree} is compared to \code{rpart}
  \citep{therneau1997introduction}, the open-source CART implementation, and
  \code{ctree} \citep{hothorn2006urp} (conditional inference trees). On 20
  evaluated datasets, including several benchmark problems from the UCI machine
  learning repository, \code{evtree} models offered at least similar and most of
  the time increased predictive accuracy.  
}

@@

<<setup, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("evtree")
@

@@

Classification and regression trees are commonly applied for exploration and modeling of complex data. They are able to handle strongly nonlinear relationships with high order interactions and different variable types. The resulting model can be interpreted as a tree structure, which gives a compact and intuitive representation. Commonly used classification and regression tree algorithms, including \textit{CART} \citep{breiman1984cart} and \textit{C4.5} \citep{quinlan1993}, use a greedy heuristic, where split rules are selected in a forward stepwise search. The data is recursively split into two groups. The split rule at each internal node is selected to maximize the homogeneity of its two child nodes, without consideration of nodes further down in the tree. Split rules selected by this top-down heuristic are only locally optimal, as splits are chosen to maximize homogeneity at the next step only. Nonetheless, the greedy heuristic is computationally efficient and often yields reasonably good results \citep{murthy1995dti}. However, for some problems, greedily induced trees can be far from the optimal solution, and a global search over the tree's parameter space can lead to much more compact and accurate models.

One possibility to build globally optimal trees is to use stochastic optimization methods like evolutionary algorithms. Evolutionary algorithms are methods that are inspired by natural Darwinian evolution. Important concepts include inheritance, mutation, and natural selection. They are population based, i.e., a whole collection of candidate solutions are processed simultaneously. The initial population is either initialized randomly, or by a problem dependent algorithm. The individuals in the populations are then iteratively modified by \textit{variation operators}. For tree algorithms variation operators can be applied to modify the tree structure and the tests that are applied in the internal nodes. Variation operators that apply to single individuals are called \textit{mutation}, while operators that merge information from different solutions are called \textit{crossover}. The survivor selection process favors individuals that perform well according to some quality criteria, which is called \textit{fitness function} or \textit{evaluation function}. In this evolutionary process the mean quality of the population increases over time \citep{baeck,eiben2003iec}.

In practice, evolutionary algorithm and other stochastic optimization methods are rarely used in decision tree induction. One reason is that they are computationally much more demanding than a forward stepwise search. Perhaps another important reason is their lack of availability in major software packages. \proglang{R} \citep{team2011r} provides several packages with tree-based methods. \pkg{rpart} \citep{therneau1997introduction} is an \proglang{R} implementations of CART. Package \pkg{party} provides two tree-based algorithms with unbiased variable selection and statistical stopping criteria. Several tree algorithms are available in \proglang{Weka} \citep{weka}. Package \pkg{RWeka} \citep{rweka} provides an interface to \proglang{Weka} for \proglang{R} users. Decision tree methods in \proglang{Weka} include implementations of C4.5 and \textit{M5} \citep{quinlan1992lcc}. However, all of these methods are based on a top-down heuristic. An exception is the \pkg{LogicReg} package, which implements logic regression. This method is applicable for regression problems with binary covariates. The resulting models of logic regression can be visualized as logic trees.

In this article, we present \pkg{evtree}, a software package that uses evolutionary methods for learning globally optimal classification and regression trees in \proglang{R}. The package is available from the Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/package=evtree}. An important difference to comparable algorithms is the survivor selection mechanism, where we use a simple $(1+1)$ selection strategy. That is, one parent solution competes with one offspring for a place in the subsequent population. As further discussed in Section~\ref{evo}, this design offers computational advantages for the application of classification and regression trees. Furthermore, we argue that the chosen selection mechanism helps to avoid premature convergence.

The remainder of this paper is structured as follows. Section~\ref{got} describes the problem of learning globally optimal decision trees and contrasts it
to the only locally optimal top-down heuristic that is utilized by recursive partitioning algorithms. In Section~\ref{desc} the \code{evtree} algorithm will
be introduced. Section~\ref{impl} addresses implementation details and gives an overview of the implemented functions. In Section~\ref{comp} \code{evtree}
is compared to the two commonly used algorithms \code{rpart} and \code{ctree}, from package \pkg{party}, which are the implementations of CART and
conditional inference trees in \proglang{R}. The comparison comprises 14~benchmark datasets, 3~real-world datasets and 3~simulated datasets. It will be shown that \code{evtree} models are often significantly more accurate compared to models from the two recursive partitioning methods. Finally, in Section~\ref{conc}, we will summarize and draw conclusions on the implementation and the performance of our algorithm.

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a function of the number of terminal nodes, without further considering the depth or the shape of trees. Then the overall parameter space $\Theta$ can be written as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the parameter space of models with $M$ terminal nodes. Let $\theta$ be a generic element of $\Theta$, then $\theta$ is fully defined by:

\begin{equation}
	 \theta=(v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ).
	\label{treeRep}
\end{equation}

The overall goal of classification and regression tree analysis can be defined as finding a set of parameters $\hat{\theta}$, such that

\begin{equation}
	\hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; loss(Y, f(X,\theta)) + comp(\theta).
	\label{generalcostfunction}
\end{equation}

Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives predictions for the output $Y$. $loss$ is a suitable loss function for the domain of $Y$. Commonly, the misclassification error for classification problems, or the squared error for regression problems, is used. $comp$ quantifies the complexity of trees and penalizes for more complex models in the tree selection process. In CART, the complexity is a non-negative function of the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all $\Theta_m$.

@@

Instead of searching all combination of $\Theta$ simultaneously, recursive partitioning algorithms only consider one split at a time. At each internal node $r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding split point $s_r$ is selected such that it minimizes the impurity in the two resulting terminal nodes. Let the parameters of a tree, constructed in $i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} = (\emptyset)$. Then, for a tree with parameters $\tilde{\theta}_{i}$, the position of the new internal node $r$, as well as the corresponding split variable $v_r$ and split point $s_r$, are chosen such that:

\begin{equation}
		\min_{v_r, s_r} \; loss(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
	\label{eq:whichsplit}
\end{equation}

CART's standard measures for node impurity are the Gini index, for classification problems and the mean squared error, for regression problems. After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by $\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$. This process of recursively selecting and executing splits continues until the homogeneity of the nodes cannot be further improved. Additionally, often additional stopping criteria are defined, e.g. minimum sample sizes in the internal nodes and terminal nodes.

In CART, and many other recursive tree methods, the tree growing step is followed by a pruning step. The idea behind pruning is to first grow a large tree without or with little restriction on the magnitude of impurity reduction. Starting from the bottom of the tree pruning then removes worthless splits. The method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and used by CART is called \textit{cost-complexity pruning}. Cost-complexity pruning iteratively prunes the internal node that:

\begin{equation}
		\min_{r=1,...,n_{M-1}} \; loss(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + comp(\tilde{\theta}_i \setminus (v_r, s_r)).
	\label{eq:pruning}
\end{equation}

The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus (\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.

@@

In contrast to CART and C4.5, conditional inference trees utilize an unbiased variable selection approach that is based on permutation testing. Bias in the split variable selection process arises from a preference towards variables with many potential split points. The conditional inference framework overcomes this problem by separating the variable and split point selection. Although this distinction influences the selection of split rules, conditional inference tree solve the same minimization problem (Equation \ref{eq:whichsplit}). The \code{ctree} implementation of conditional inference trees does not employ pruning, but uses a statistical stopping criterion.

\subsection{An illustration showing the limitations of locally optimal decision trees}

@@

This section continues with a sample session where \code{evtree} is used, in conjunction with \code{rpart} and \code{ctree}, to analyze the \textit{Bookbinder's Book Club} data. In this session, all trees are constrained to have a minimum number of $10$ observations per terminal node. To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$. To provide uniform visualizations and predictions of the fitted models the \code{as.party()} function is used to convert \pkg{rpart} objects into objects of class \pkg{party}. The sample session starts with the construction and visualization of trees using the \code{rpart} and \code{ctree} methods.
%
<<rpart-ctree>>=
data("BBBClub", package = "evtree")
library("rpart")
set.seed(1090)
rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
ct <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
plot(rp)
plot(ct)
@
%
The resulting trees are displayed in Figure~\ref{fig:example_rpart_ctree}.
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
<<rp-plot, fig=TRUE, echo=FALSE, width=12, height=6.8>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
<<ct-plot, fig=TRUE, echo=FALSE,  width=12, height=6.8>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees constructed by \code{rpart} (upper panel) and \code{ctree} (lower panel). The target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the Bookbinder's Book Club; \code{gender} the customers gender.
}
\end{figure}
%
With the objective of building a smaller, but at least equally accurate tree, \code{evtree} is constrained to a maximum tree depth of $2$. The resulting tree is displayed in Figure~\ref{fig:example_evtree}. Alternatively, the tree can be examined by the textual output provided by the \code{print()} function.
%
<<evtree-plot>>=
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
plot(ev)
print(ev)
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
<<iris-plot, fig=TRUE, echo=FALSE, width=12, height=7>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree constructed by \code{evtree}. The target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase.}
\end{figure}
%
While the \code{rpart} model has $5$ terminal nodes and a tree depth of $4$, the tree constructed by \code{ctree} has $7$ terminal nodes and a tree depth of $3$. With a tree depth of $2$ and $4$ terminal nodes the \code{evtree} model is much simpler. 
%
<<evtree-performance>>=
mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
@
The accuracy of the three models are similar. With a misclassification rate of $24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model ($23.8\%$) and better than the \code{ctree} model ($24.8\%$). 
%
<<evtree-structure>>=
ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
  ctree  = predict(ct), observed = BBBClub$choice))

@@

ciplot(acc_rpart, xlim = xlim1, main = "rpart", ylab = TRUE, xlab = "")
ciplot(com_rpart, xlim = xlim2, main = "",      ylab = FALSE, xlab = "")
ciplot(acc_ctree, xlim = xlim1, main = "ctree", ylab = TRUE,
  xlab = "relative difference in predictive accuracy (%)")
ciplot(com_ctree, xlim = xlim2, main = "",      ylab = FALSE,
  xlab = "relative difference in complexity (%)")
@
\caption{\label{fig:benchmark} Performance comparison of \code{evtree} vs.\ \code{rpart} (upper panels) and \code{ctree} (lower panels).
  Prediction error (left panels) is compared by the relative difference of the misclassification rate or the mean-squared error.
  The complexity (right panels) is compared by the relative difference of the number of terminal nodes.}
\end{figure}

>>>>

\Abstract{
  Commonly used classification and regression tree methods like the CART
  algorithm are recursive partitioning methods that build the model in a forward
  stepwise search. Although this approach is known to be an efficient heuristic,
  the results of recursive tree methods are only locally optimal, as splits are
  chosen to maximize homogeneity at the next step only. An alternative way to
  search over the parameter space of trees is to use global optimization methods
  like evolutionary algorithms. This paper describes the \pkg{evtree} package,
  which implements an evolutionary algorithm for learning globally optimal
  classification and regression trees in \proglang{R}. Computationally intensive
  tasks are fully computed in \proglang{C++} while the \pkg{partykit}
  \citep{hothorn2009partykit} package is leveraged for representing the resulting
  trees in \proglang{R}, providing unified infrastructure for summaries,
  visualizations, and predictions. \code{evtree} is compared to \code{rpart}
  \citep{therneau1997introduction}, the open-source CART implementation, and
  conditional inference trees \citep[\code{ctree},][]{hothorn2006urp}. The usefulness
  of \code{evtree} is illustrated in a textbook customer classification task
  and a benchmark study of predictive accuracy in which \code{evtree}
  achieved at least similar and most of the time better results compared to
  the recursive algorithms \code{rpart} and \code{ctree}.  
}

@@

<<setup, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("rpart")
library("evtree")
data("BBBClub", package = "evtree")
cache <- TRUE
@

@@

Classification and regression trees are commonly applied for exploration and
modeling of complex data. They are able to handle strongly nonlinear
relationships with high order interactions and different variable types. The
resulting model can be interpreted as a tree structure providing a compact
and intuitive representation. Commonly used classification and regression tree
algorithms, including \textit{CART} \citep{breiman1984cart} and \textit{C4.5}
\citep{quinlan1993}, use a greedy heuristic, where split rules are selected in a
forward stepwise search for recursively partitioning the data into groups. The
split rule at each internal node is selected to maximize the homogeneity of its
child nodes, without consideration of nodes further down the tree, hence yielding
only locally optimal trees. Nonetheless, the
greedy heuristic is computationally efficient and often yields reasonably good
results \citep{murthy1995dti}. However, for some problems, greedily induced
trees can be far from the optimal solution, and a global search over the tree's
parameter space can lead to much more compact and accurate models.

The main challenge in growing globally optimal trees is that the search
space is typically huge rendering full-grid searches computationally infeasible.
One possibility to solve this problem is to use stochastic
optimization methods like evolutionary algorithms. 
In practice, however, such stochastic methods
are rarely used in decision tree induction. One reason is probably that they are
computationally much more demanding than a recursive forward search but another one is
likely to be the lack of availability in major software packages.
In particular, while there are several packages for \proglang{R} \citep{team2011r} providing
forward-search tree algorithms, there is only little support for globally
optimal trees. The former group of packages includes (among others)
\pkg{rpart} \citep{therneau1997introduction}, the open-source
implementation of the CART algorithm; \pkg{party}, containing two tree algorithms
with unbiased variable selection and statistical stopping criteria
\citep{hothorn2006urp,zeileis2008mob}; and \pkg{RWeka} \citep{rweka}, the
\proglang{R} interface to \pkg{Weka} \citep{weka} with open-source implementations
of tree algorithms such as C4.5 or \textit{M5} \citep{quinlan1992lcc}.
A notable exception is the
\pkg{LogicReg} package \citep{logicreg} for logic regression, an algorithm for globally optimal
trees based on binary covariates only  and using simulated annealing. See
\cite{hothorn2011ctv} for an overview of further recursive partitioning packages
for \proglang{R}.

To fill this gap, we introduce a new \proglang{R} package \pkg{evtree}, 
available from the Comprehensive \proglang{R} Archive Network at
\url{http://CRAN.R-project.org/package=evtree}, providing evolutionary
methods for learning globally optimal classification and regression trees.
Generally speaking, evolutionary algorithms are inspired by natural Darwinian
evolution employing concepts such as inheritance, mutation, and natural selection.
They are population-based, i.e., a whole collection of candidate solutions -- trees in this application -- is
processed simultaneously and iteratively modified by \textit{variation operators}
called \textit{mutation} (applied to single solutions) and \textit{crossover}
(merging different solutions). Finally, a survivor selection process favors
solutions that perform well according to some quality criterion, usually
called \textit{fitness function} or \textit{evaluation function}. In this
evolutionary process the mean quality of the population increases over time
\citep{baeck,eiben2003iec}.  
In the case of learning decision trees, this means that the variation
operators can be applied to modify the tree structure (e.g., number of splits, splitting
variables, and corresponding splitpoints etc.) in order to optimize a
fitness functions such as the misclassificaion or error rate penalized by
the complexity of the tree.
A notable difference to comparable algorithms is the survivor selection
mechanism where it is important to avoid premature convergence. In the
following, we use a simple $(1+1)$ selection strategy (i.e., one parent
solution competes with one offspring for a place in the population)
which can be argued to offer computational
advantages for the application to classification and regression trees.

The remainder of this paper is structured as follows: Section~\ref{got}
describes the problem of learning globally optimal decision trees and contrasts
it to the locally optimal top-down heuristic that is utilized by recursive
partitioning algorithms. Section~\ref{desc} introduces the \code{evtree} algorithm
before Section~\ref{impl} addresses implementation details along with an
overview of the implemented functions. A benchmark comparison -- comprising
14~benchmark datasets, 3 real-world datasets, and 3 simulated scenarios -- is carried out
in Section~\ref{comp}, showing that the predictive performance of \code{evtree}
is often significantly better compared to commonly used algorithms \code{rpart}
and \code{ctree} (from \pkg{party} package). Finally, Section~\ref{conc} gives
concluding remarks about the implementation and the performance of the new
algorithm.

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. Then the overall parameter space $\Theta$ can be written
as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the
parameter space of models with $M$ terminal nodes. Let $\theta$ be a generic
element of $\Theta$, then $\theta$ is fully defined by:
%
\begin{equation}
	 \theta=(v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ).
	\label{treeRep}
\end{equation}
%
The overall goal of classification and regression tree analysis can be defined
as finding a set of parameters $\hat{\theta}$, such that
%
\begin{equation}
	\hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; loss(Y, f(X,\theta)) + comp(\theta).
	\label{generalcostfunction}
\end{equation}
%
Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives
predictions for the output $Y$. $loss$ is a suitable loss function for the
domain of $Y$. Commonly, the misclassification error for classification
problems, or the squared error for regression problems, is used. $comp$
quantifies the complexity of trees and penalizes for more complex models in the
tree selection process. In CART, the complexity is a non-negative function of
the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all
$\Theta_m$.

@@

Instead of searching all combination of $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding
split point $s_r$ is selected such that it minimizes the impurity in the two
resulting terminal nodes. Let the parameters of a tree, constructed in
$i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} =
(\emptyset)$. Then, for a tree with parameters $\tilde{\theta}_{i}$, the
position of the new internal node $r$, as well as the corresponding split
variable $v_r$ and split point $s_r$, are chosen such that:
%
\begin{equation}
		\min_{v_r, s_r} \; loss(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
	\label{eq:whichsplit}
\end{equation}
%
CART's standard measures for node impurity are the Gini index, for
classification problems and the mean squared error, for regression problems.
After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by
$\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$. This
process of recursively selecting and executing splits continues until the
homogeneity of the nodes cannot be further improved. Additionally, often
additional stopping criteria are defined, e.g. minimum sample sizes in the
internal nodes and terminal nodes.

In CART, and many other recursive tree methods, the tree growing step is
followed by a pruning step. The idea behind pruning is to first grow a large
tree without or with little restriction on the magnitude of impurity reduction.
Starting from the bottom of the tree pruning then removes worthless splits. The
method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and
used by CART is called \textit{cost-complexity pruning}. Cost-complexity pruning
iteratively prunes the internal node that:
%
\begin{equation}
		\min_{r=1,...,n_{M-1}} \; loss(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + comp(\tilde{\theta}_i \setminus (v_r, s_r)).
	\label{eq:pruning}
\end{equation}
%
The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus
(\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.

@@

In contrast to CART and C4.5, conditional inference trees utilize an unbiased
variable selection approach that is based on permutation testing. Bias in the
split variable selection process arises from a preference towards variables with
many potential split points. The conditional inference framework overcomes this
problem by separating the variable and split point selection. Although this
distinction influences the selection of split rules, conditional inference tree
solve the same minimization problem (Equation~\ref{eq:whichsplit}). The
\code{ctree} implementation of conditional inference trees does not employ
pruning, but uses a statistical stopping criterion.

\subsection{An illustration of the limitations of locally optimal decision trees}

@@

This section continues with a sample session where \code{evtree} is used, in conjunction with \code{rpart} and \code{ctree}, to analyze the \textit{Bookbinder's Book Club} data. In this session, all trees are constrained to have a minimum number of $10$ observations per terminal node. To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$. To provide uniform visualizations and predictions of the fitted models the \code{as.party()} function is used to convert \pkg{rpart} objects into objects of class \pkg{party}. The sample session starts with the construction and visualization of trees using the \code{rpart} and \code{ctree} methods.
%
<<BBBClub-rpart-ctree, echo=TRUE, eval=FALSE>>=
data("BBBClub", package = "evtree")
library("rpart")
rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
ct <- ctree(choice ~ ., data = BBBClub, minbucket = 10, mincrit = 0.99)
plot(rp)
plot(ct)
@
%
The resulting trees are displayed in Figure~\ref{fig:example_rpart_ctree}.
%
With the objective of building a smaller, but at least equally accurate tree,
\code{evtree} is constrained to a maximum tree depth of $2$.
%
<<BBBClub-evtree, echo=TRUE, eval=FALSE>>=
set.seed(1090)
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
@
%
<<BBBClub-cache, echo=FALSE, eval=TRUE>>=
if(cache & file.exists("BBBClub-trees.rda")) {
load("BBBClub-trees.rda")
} else {
<<BBBClub-rpart-ctree>>
<<BBBClub-evtree>>
if(cache) {
  save(rp, ct, ev, file = "BBBClub-trees.rda")
} else {
  if(file.exists("BBBClub-trees.rda")) file.remove("BBBClub-trees.rda")
}
}
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[p!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<BBBClub-rpart-plot, fig=TRUE, echo=FALSE, width=8.1, height=7>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
\setkeys{Gin}{width=1\textwidth}
<<BBBClub-ctree-plot, fig=TRUE, echo=FALSE,  width=11.4, height=5.9>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees constructed by \code{rpart} (upper panel) and \code{ctree} (lower panel). The target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the Bookbinder's Book Club; \code{gender} the customers gender.
}
\end{figure}
%
The resulting tree is displayed in the printed output below and visualized in
Figure~\ref{fig:example_evtree}.
%
<<BBBClub-evtree-display>>=
plot(ev)
ev
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<BBBClub-evtree-plot, fig=TRUE, echo=FALSE, width=7.5, height=4.9>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree constructed by \code{evtree}. The target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase.}
\end{figure}
%
While the \code{rpart} model has \Sexpr{width(rp)}~terminal nodes and a tree depth
of~\Sexpr{depth(rp)}, the tree constructed by \code{ctree} has
\Sexpr{width(ct)}~terminal nodes and a tree depth of~\Sexpr{depth(ct)}.
With a tree depth of~\Sexpr{depth(ev)} and \Sexpr{width(ev)}~terminal nodes
the \code{evtree} model is much simpler. 
%
<<evtree-performance>>=
mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
@
The accuracy of the three models are similar. With a misclassification rate of
$24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model
($23.8\%$) and better than the \code{ctree} model ($24.8\%$). 
%
<<evtree-structure>>=
ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
  ctree  = predict(ct), observed = BBBClub$choice))

@@

ciplot(acc_rpart, xlim = xlim1, main = "rpart", ylab = TRUE, xlab = "")
ciplot(com_rpart, xlim = xlim2, main = "",      ylab = FALSE, xlab = "")
ciplot(acc_ctree, xlim = xlim1, main = "ctree", ylab = TRUE,
  xlab = "relative difference in predictive accuracy (%)")
ciplot(com_ctree, xlim = xlim2, main = "",      ylab = FALSE,
  xlab = "relative difference in complexity (%)")
@
\caption{\label{fig:benchmark} Performance comparison of \code{evtree} vs.\ \code{rpart} (upper panels) and \code{evtree} vs.\ \code{ctree} (lower panels).
  Prediction error (left panels) is compared by the relative difference of the misclassification rate or the mean-squared error.
  The complexity (right panels) is compared by the relative difference of the number of terminal nodes.}
\end{figure}
